Competing Constructions for German Measure Noun Phrases: from Usage Data to Experimental Validation

### 2017-04-02 Latex/article.tex
fixed a serious flaw in theory

In this section, I briefly motivate the choice of statistical models which I use later in Section~\ref{sec:corpusstudies}.
In the spirit of the careful and methodologically sound approach to cognitively oriented corpus linguistics laid out in Section~\ref{sec:cogocl}, authors should always provide at least a minimal motivation as part of the standard protocol, especially if there are alternative methods.

@@

In the case at hand, three different syntactic configurations need to be distinguished \wrt case assignment inside German measure noun phrases.
If the kind noun forms an NP with a determiner, the construction resembles (and is usually called) a \textit{pseudo-partitive} (on partitives and pseudo-partitives see, e.\,g., \citealp{Barker1998,Selkirk1977,Stickney2007,Vos1999}; for a recent application of the terminology to German, see \citealp{Gerstenberger2015}).%
\footnote{If the kind noun is definite, the construction instantiates a true partitive.
Whereas partitives are constructions denoting a proper part-of relation as in \textit{a sip of the wine}, pseudo partitives -- albeit syntactically similar and diachronically related to partitives in many languages -- merely denote quantities and contain indefinite kind nouns as in \textit{a sip of wine}.
%In the literature on German, some authors incorrectly call the pseudo-partitive a \textit{partitive} \cite{Hentschel1993} while some realise the difference and at least mention it \citep{Eschenbach1994,GallmannLindauer1994,Loebel1989,Zimmer2015}.
}
Here, the kind noun is in the genitive, and I refer to the construction in (\ref{ex:intro:pseudopartitive1}) as the \textit{Pseudo-partitive Genitive Construction} (PGC).

@@

The distribution of the case patterns in the NAC (case identity between the measure noun and the kind noun) and in the PGC (genitive on the kind noun, regardless of the case of the measure noun) depending on the structure of the kind NP are summarised in Table~\ref{tab:constructions}.
I call the narrow apposition construction with a bare kind noun the \NACb, the partitive genitive with a determiner in the kind noun phrase the \PGCd.
I call the alternating variants with an adjective but no determiner in the kind noun phrase \NACa, and \PGCa, respectively.

@@

\begin{enumerate}[i.]
  \item{\label{it:intro:idiot} alleged alternatives to the NAC and the PGC}
  \item{\label{it:intro:datsg} alternative \textit{weak} forms of dative singular neuter adjectives}
%  \item{\label{it:intro:femsg} case syncretism in feminine NPs}
  \item{\label{it:intro:plurl} similar constructions with plural/collective kind nouns}
  \item{\label{it:intro:noifl} grammaticalised non-inflected measure nouns}
  \item{\label{it:intro:other} alternative constructions for expressing quantities}
\end{enumerate}

@@

First of all, (\ref{it:intro:idiot}) refers to claims found in some grammars that a generic nominative, accusative, and even dative on the kind noun are used instead of the genitive or case agreement (overviews in \citealp{Hentschel1993,Zimmer2015}).
It was shown empirically in \cite{Hentschel1993} that such variants are de facto not acceptable.
Also, in my corpus sample, they simply did not occur.
Even if they are accepted by some speakers, their extremely low frequency makes it virtually impossible to study them using either corpus methods (Section~\ref{sec:corpusstudies}) or experimental approaches (Section~\ref{sec:externalvalidation}), and I consequently ignore them.

@@

Finally, (\ref{it:intro:other}) suggests that there are alternative ways of expressing similar quantificational meanings.
In the variationist tradition, which is strongly influenced by Labovian sociolinguistics, the \textit{principle of accountability} would dictate that proper studies should examine a variationist \textit{variable}, \ie\ all different \textit{ways of saying the same thing} (\citealp{Labov1966}, \citealp{Labov1969}, for an overview see \citealp{Tagliamonte2012}).
In the case at hand, the variable might be something like \textit{measuring substances and collections}.
I argue that it is fully justified to focus narrowly on \NACa\ and \PGCa\ with their well-defined morph-syntactic properties, mostly because the alternative constructions are not used in the same range of contexts.
Two major alternatives might be considered.
First of all, the analytic (pseudo-)partitive with \textit{von} `of' is only available as an alternative to the PGC if the kind noun phrase contains a (definite or indefinite) determiner as in (\ref{ex:analyticalpartitive}).

@@

This section briefly reviews existing analyses of the \PGCa vs.\ \NACa\ alternation and related issues.
I also develop my own analysis and the appropriate hypotheses for the empirical studies presented in Sections~\ref{sec:corpusstudies} and \ref{sec:externalvalidation}.
I discuss four main factors that might influence the alternation:
first, the different syntactic structures of the variants \NACa\ and \PGCa\ in relation to their syntactic prototypes \NACb\ and \PGCd;
second, degrees of grammaticalisation of the measure noun associated with the two prototypes;
third, a semantic prototypicality effect of the degrees of grammaticalisation (plurality);
fourth, preferences in different registers.

The first aspect is related to analyses of the morpho-syntax of German MNPs.
A major interest in descriptive and generative linguistics is to find a plausible analysis of the hierarchical syntactic structure of the constructions in question.
The first question is which noun constitutes the head of the whole construction.
This was answered quite clearly by \cite{Loebel1986} already (see also \citealp[213]{Eschenbach1994}, and \citealp[16]{GallmannLindauer1994}).
Subject-verb agreement is always realised on the measure noun, and we can therefore assume that the measure noun is the head.
However, the MNP-internal structure has more interesting cues to offer if the strong\slash weak inflection patterns of adjectives are taken into account.
In NPs with a strongly inflected determiner, attributive adjectives inflect according to the massively syncretistic \textit{weak} pattern.
%\footnote{The weak pattern has only two suffixes.
%The structural cases (nominative and accusative) in the singular are marked with \textit{-e}, the oblique cases (dative and genitive) as well as the entire plural are marked with \textit{-en}.
%The only exception to this rule is the masculine accusative singular, which is also marked with \textit{-en}.}
If there is no determiner (as is the case in the alternating constructions), attributive adjectives inflect like determiners themselves, however.
This is called the \textit{strong} inflectional pattern.
Thus, the adjectives in the \NACa\ and the \PGCa\ have properties of adjectives as well as determiners.
They are lexical adjectives and function as attributive modifiers.
On the other hand, they are inflected like determiners, and they are the leftmost element in the NP, which is typical of determiners.
This unusual double nature of adjectives in NPs without determiners leads to a plausible probabilistic interpretation of the pattern shown in Table~\ref{tab:constructions}.
Whenever speakers classify the adjective in the kind noun phrase more as a determiner, they have to use the \PGCa\ because, if there is a determiner, the PGC (just like in the case of the \PGCd) is the only option.
When they classify the adjective more as an adjective, the kind NP has no determiner, and they have to use the \NACa\ (like in the case of the \NACb).%
\footnote{While the generative analysis presented in \cite{Bhatt1990} cannot properly deal with probabilistic effects, Bhatt comes close to this interpretation by analysing the kind NP in the GPC as a DP and in the NAC as an NP.}
In a framework that assumes similarity-based probabilistic classification, such as Prototype Theory, we can regard the \PGCd\ as instantiating the PGC prototype and the \NACb\ as the NAC prototype.
The choice between \PGCa\ and \NACa\ is then expected to be influenced by the degree of similarity to either of these prototypes.
I will argue below that the prototypes differ mainly in that they attract different (classes of) measure nouns more or less strongly, mostly because of degrees of grammaticalisation.
However, if lemmas are attracted with different strengths by the two prototypes, then we can expect a baseline attraction effect:
nouns which occur more often in the \PGCd\ should favour the \PGCa, and nouns which occur more often in the \NACb\ should favour the \NACa.
Put differently, a noun's relative frequencies in the prototypical non-alternating constructions should be indicative of same noun's preference to appear in the two alternating variants.
This baseline attraction effect will be quantified in the corpus study Section~\ref{sec:corpusstudies}.

@@

Furthermore, the grammaticalisation path as described above leads from NPs denoting individuated objects standing in a \textit{part-of} relation to a construction with a more diffuse \textit{quantity-of} relation.
The best evidence for the relevance of this factor in an ongoing grammaticalisation process is that plural kind nouns almost always (but not exclusively) occur in the \PGCa\ and virtually never in the \NACa\ (see Section~\ref{sec:descriptive}).
More indirectly, however, plurality of the measure noun could also favour the \PGCa\ over the \NACa, and exact quantification in the form of cardinals even more so.%
\footnote{Without offering a theoretical analysis, such a preference was observed empirically by \cite{Hentschel1993}.}
There are highly illustrative examples with measure nouns that have virtually developed two variants, one of which grammaticalised to a much higher degree, \eg\ \textit{Haufen} `heap\slash lot'.
The phrase \textit{ein Haufen französische Watte} `a heap\slash lot of French cotton' (\NACa) has two uses.
It can be used to denote an unspecific amount of cotton in the sense of \textit{a lot of French cotton} in situations when the exact quantity of cotton is not of interest.
Under the appropriate circumstances, it could also be used to denote a heap-shaped, potentially highly specific quantity of cotton, for example when a cotton merchant always sells her French cotton in \textit{heaps} weighing two stone each.
The former use, however, is highly restricted for \textit{zwei Haufen französischer Watte} `two heaps of French cotton' (\PGCa) -- not `two lots of French cotton'.
Such an NP is less likely to be used in situations where the exact quantity of cotton is irrelevant, even if there are actually two heaps of cotton.
In other words, a precisely quantified measure noun leads to a higher individuation (the creation of discrete portions) of the substance denoted by the kind noun, which is more appropriate for the PGC prototype.
The statistical models in Section~\ref{sec:corpusstudies} will be designed to detect such an effect.
%\footnote{In the extreme, some measure nouns cannot be pluralised, such as \textit{eine Menge} `a lot of'.
%Also note that the presentation of the examples in NAC and PGC is suggestive.
%Native speakers will confirm that both NPs sound highly acceptable in both GPC and NAC.
%However, it was made clear from the beginning and throughout that the preferences are indeed subtle.}

Finally, it has been iterated very often that the PGC is more typical of higher registers or even exclusive to written language (see \citealp[320--323]{Hentschel1993}, where she also refers to descriptive grammars).
This is not surprising in as much as the genitive -- an intrinsic part of the PGC -- is generally underrepresented in colloquial and vernacular variants of German.
Under an integral view, such preferences can be to be part of the construction prototypes.
In the corpus study in Section~\ref{sec:corpusstudies}, register effects will therefore be modelled, even if only using some very simple proxies.

@@

\begin{enumerate}[i.]
  \item The relative frequencies with which measure noun lemmas and kind noun lemmas appear in the prototypical (non-alternating) \PGCd\ and \NACb\ are positively correlated with the relative frequency the alternating \PGCa\ and the \NACa\ are chosen.
  \item (Classes of) more strongly grammaticalised measure nouns favour the NAC.
  \item Plural measure nouns and measure nouns quantified by cardinals favour the PGC variant.
  \item The NAC is associated more strongly with higher registers.
\end{enumerate}

@@

For the present study, I used the German \textit{Corpus from the Web} (COW) in its 2014 version DECOW14A (\citealp{SchaeferBildhauer2012full,Schaefer2015b} and \citealp{BiemannEa2013,SchaeferBildhauer2013} for overviews of web corpora in general and the methodology of their construction), which contains almost 21 billion tokens.%
\footnote{The corpora are made available for free at \url{https://www.webcorpora.org}.
At the time of this writing, a newer 2016 version DECOW16 had already been released.}
I chose this corpus for two main reasons.
First, the external validity of any study is increased through a higher heterogeneity of the sample \citep[30]{MaxwellDelaney2004}, and the DECOW corpus has clearly a much more heterogeneous composition compared to the only other very large corpus of German, the DeReKo \citep{KupietzEa2010} of the Institute for the German Language (IDS), which contains almost exclusively newspaper texts.%
\footnote{It was shown in \cite{W16-2601} that, for example, the spread of topics is much smaller in DeReKo compared to DECOW.}
Second, it was already mentioned that normative grammars often adopt clear positions regarding the grammaticality of either the \NACa\ or the \PGCa.
Thus, newspaper text or any other text that conforms strongly to normative grammars might not represent the alternation phenomenon fully (and without bias) because authors and proof-readers might favour one alternative or the other.
Web corpora, on the other hand, contain at least some amount of non-standard language from forums and similar sources.
For these or similar reasons, COW corpora have been used in a number of peer-reviewed publications, for example \cite{VanGoethemHiligsmann2014,VanGoethemHuening2015,MuellerS2014,Schaefer2016c,SchaeferSayatz2014,SchaeferSayatz2016,Zimmer2015}. 
Therefore, DECOW can be considered the obvious choice for this study.

@@

I used exactly the same stimuli as in the forced choice experiments.
Each participant read both the $16$ sentences with the variant predicted by the corpus model and the $16$ modified sentences with the variant that the corpus model does not predict.%
\footnote{Notice that lemmas and their frequencies as well as lemma classes are part of the corpus-based GLMM, and there was no additional controlling of lemma frequencies, etc.}
To avoid repetition effects, the stimuli for each participant were separated into two blocks of $16$ targets and $33$ fillers per block.
It was made sure that from each target sentence pair, one sentence was assigned to the first block, and the other sentence to the other block.
The assignment of members of the individual sentence pairs to one of the blocks was randomised for each participant individually, and so was the order within the block.
The sentences from each pair of variants were kept as far apart as possible.
The fillers also came in pairs such that the second block exclusively contained sentences to which participants had been exposed in the first block in slightly modified form.
In total, each participant read $98$ sentences.
After each sentence, participants had to answer simple (non-metalinguistic) yes-no questions about the previous sentence as distractors.
The distractor questions were different ones in the first and the second block.
There were $38$ (different) participants recruited in exactly the same manner as for the experiment reported in Section~\ref{sec:exp:fc}.
The experiment was conducted using \textit{PsychoPy}.

The reading times were residualised per speaker based on the reading times of all words (not just the targets) by that speaker.
The adjective and kind noun (\ie\ the constituents bearing the critical case markers) were used as the target region, such as the bracketed words in the example \textit{zwei Gläser} [\textit{sprudelndes Wasser}] `two glasses of sparkling water'.
Outliers farther than 2 inter-quartile ranges from the mean logarithmised residual time were reamoved ($64$ data points), resulting in a total number of $n=1,152$ observations.
An LMM was specified with the logarithmised residual reading times as the response variable, and the probabilities derived from the corpus GLMM (\textit{Modelprediction}) as the main regressor.
Since the corpus GLMM predicts the probability of the \PGCa (vs.\ the \NACa), it is expected that reading times are positively correlated (longer reading times) with the probability if the stimulus actually contains the \NACa, and negatively correlated (shorter reading times) if the stimulus actually contains the \PGCa.
To account for this, an interaction between \textit{Modelprediction} and \textit{Construction} (levels \textit{PGCa} and \textit{NACa}) was added to the model.
Furthermore, the position of the sentence in the individual experiment run was included as a fixed effect to control for increasing reading speed during experiment runs.
Random intercepts were specified for \textit{Participant} and \textit{Item} (the $16$ sentence pairs are one \textit{Item} each).%
\footnote{I tried random slopes in order to keep the random effect structure maximal \citep{BarrEa2013}, but it was impossible to get the algorithm to converge.}

@@

While the overall model quality is acceptable, the relevant effect is estimated with some uncertainty, which is also visible in the effect plot in Figure~\ref{fig:spr:effects}.
While the estimate for the sentences with \NACa\ is obviously very imprecise, there is a clearer tendency in the sentences with \PGCa, which is also confirmed by the `significant' results from the bootstrapped confidence intervals (see Table~\ref{tab:exp:spr}) and from the PB test reported above.
The \PGCa\ brings about an increased reading time ($\beta_{\text{ConstructionPGCa}}=0.062$), which is plausible because it is the much rarer construction (see Section~\ref{sec:corpusstudies}).
However, if it occurs in a prototypical situation, reading times drop ($\beta_{\text{ConstructionPGCa:Modelprediction}}=-0.132$).
This is noticeable in the downward slope of curve in the right panel of Figure~\ref{fig:spr:effects}.
This fits into the general picture in as much as the construction with the lower frequency might be developing towards a more sharply defined prototype.%
\footnote{In this context, it should be remembered from Section~\ref{sec:gettingdata} that even the \PGCd\ is much rarer that the \NACb\ ($17,252$ vs.\ $315,635$ occurrences in the auxiliary corpus samples).}
Conversely, the \NACa\ (like the NAC in general) might be the highly frequent default which does not incur reading time penalty, even if it is not the optimal choice in the given context and with the given lexical material.
This interpretation will be considered, among other things, in the next and final section.

>>>>

In this section, I briefly motivate the choice of statistical models which I use later in Section~\ref{sec:corpusstudies}.

@@

In the case at hand, three different syntactic configurations need to be distinguished \wrt case assignment inside German measure noun phrases.
If the kind noun forms an NP with a determiner, the construction resembles (and is usually called) a \textit{pseudo-partitive} (on partitives and pseudo-partitives see, e.\,g., \citealp{Barker1998,Selkirk1977,Stickney2007,Vos1999}; for a recent application of the terminology to German, see \citealp{Gerstenberger2015}).%
\footnote{If the kind noun is definite, the construction instantiates a true partitive.
Whereas partitives are constructions denoting a proper part-of relation as in \textit{a sip of the wine}, pseudo partitives -- albeit syntactically similar and diachronically related to partitives in many languages -- merely denote quantities and contain indefinite kind nouns as in \textit{a sip of wine}.%
In the literature on German, some authors incorrectly call the pseudo-partitive a \textit{partitive} \cite{Hentschel1993} while some realise the difference and at least mention it \citep{Eschenbach1994,GallmannLindauer1994,Loebel1989,Zimmer2015}.
}
Here, the kind noun is in the genitive, and I refer to the construction in (\ref{ex:intro:pseudopartitive1}) as the \textit{Pseudo-partitive Genitive Construction} (PGC).

@@

The distribution of the case patterns in the NAC (case identity between the measure noun and the kind noun) and in the PGC (genitive on the kind noun, regardless of the case of the measure noun) depending on the structure of the kind NP are summarised in Table~\ref{tab:constructions}.
I call the narrow apposition construction with a bare kind noun the \NACb, the partitive genitive with a determiner in the kind noun phrase the \PGCd.
I call the alternating variants with an adjective but no determiner in the kind noun phrase \NACa, and \PGCa, respectively.
This paper is about the middle column of Table~\ref{tab:constructions}, \ie, the syntactic configuration in which two different case patterns are acceptable.

@@

\begin{enumerate}[i.]
  \item{\label{it:intro:idiot} alleged alternatives to the \NACa\ and the \PGCa}
  \item{\label{it:intro:datsg} alternative \textit{weak} forms of dative singular neuter adjectives}
%  \item{\label{it:intro:femsg} case syncretism in feminine NPs}
  \item{\label{it:intro:plurl} similar constructions with plural/collective kind nouns}
  \item{\label{it:intro:noifl} grammaticalised non-inflected measure nouns}
  \item{\label{it:intro:other} alternative constructions for expressing quantities}
\end{enumerate}

@@

First of all, (\ref{it:intro:idiot}) refers to claims found in some grammars that a generic nominative, accusative, and even dative on the kind noun are used instead of the genitive (\PGCa) or case agreement (\NACa).
Overviews can be found in \citealp{Hentschel1993,Zimmer2015}.
It was shown empirically in \cite{Hentschel1993} that such variants are de facto not acceptable.
Also, in my corpus sample, they simply did not occur.
Even if they are accepted by some speakers, their extremely low frequency makes it virtually impossible to study them using either corpus methods (Section~\ref{sec:corpusstudies}) or experimental approaches (Section~\ref{sec:externalvalidation}), and I consequently ignore them.

@@

Finally, (\ref{it:intro:other}) suggests that there are alternative ways of expressing similar quantificational meanings.
In the variationist tradition, which is strongly influenced by Labovian sociolinguistics, the \textit{principle of accountability} would dictate that proper studies should examine a variationist \textit{variable}, \ie\ all different \textit{ways of saying the same thing} (\citealp{Labov1966}, \citealp{Labov1969}, for an overview see \citealp{Tagliamonte2012}).
In the case at hand, the variable might be something like \textit{measuring quantities of substances and collections}.
I argue that it is fully justified to focus narrowly on \NACa\ and \PGCa\ with their well-defined morph-syntactic properties, mostly because the alternative constructions are not used in the same range of contexts.
Two major alternatives might be considered.
First of all, the analytic (pseudo-)partitive with \textit{von} `of' is only available as an alternative to the PGC if the kind noun phrase contains a (definite or indefinite) determiner as in (\ref{ex:analyticalpartitive}).

@@

This section briefly reviews existing analyses of the \PGCa vs.\ \NACa\ alternation and related issues.
I also develop my own analysis and the appropriate hypotheses for the empirical studies presented in Sections~\ref{sec:corpusstudies} and \ref{sec:externalvalidation}.
I discuss four main aspects:
first, the different syntactic structures of the variants \NACa\ and \PGCa\ in relation to their syntactic prototypes \NACb\ and \PGCd;
second, degrees of grammaticalisation of the measure noun associated with the two prototypes;
third, a semantic prototypicality effect of the degrees of grammaticalisation (preference to occur with cardinals);
fourth, preferences in different registers.

% MORPHO-SYNTAX AND PROTOTYPES

My whole analysis is based on the idea that the syntactic structure [N\Sub{1}~[A~N\Sub{2}]\Sub{NP\Sub{2}}]\Sub{NP\Sub{1}} as instantiated in the \NACa\ and the \PGCa\ is ambiguous.
To show this, the first question is which noun constitutes the head of the whole construction.
This was answered quite clearly by \cite{Loebel1986} already (see also \citealp[213]{Eschenbach1994}, and \citealp[16]{GallmannLindauer1994}).
Subject-verb agreement is always realised on the measure noun, and we can therefore assume that the measure noun is the head.
However, the MNP-internal structure has more interesting cues to offer if the strong\slash weak inflection patterns of adjectives are taken into account.
In NPs with a strongly inflected determiner, attributive adjectives inflect according to the massively syncretistic \textit{weak} pattern.
%\footnote{The weak pattern has only two suffixes.
%The structural cases (nominative and accusative) in the singular are marked with \textit{-e}, the oblique cases (dative and genitive) as well as the entire plural are marked with \textit{-en}.
%The only exception to this rule is the masculine accusative singular, which is also marked with \textit{-en}.}
If there is no determiner (as is the case in the alternating constructions), attributive adjectives inflect like determiners themselves, however.
This is called the \textit{strong} inflectional pattern.
Thus, the adjectives in the \NACa\ and the \PGCa\ have properties of adjectives as well as determiners.
They are lexical adjectives and function as attributive modifiers.
On the other hand, they are inflected like determiners, and they are the leftmost element in the NP, which is typical of determiners.
This unusual double nature of adjectives in NPs without determiners leads to a plausible probabilistic interpretation of the pattern shown in Table~\ref{tab:constructions}.
Whenever speakers classify the adjective in the kind noun phrase more as a determiner, they have to use the \PGCa\ because, if there is a determiner, the PGC (just like in the case of the \PGCd) is the only option.
When they classify the adjective more as an adjective, the kind NP has no determiner, and they have to use the \NACa\ (like in the case of the \NACb).%
\footnote{While the generative analysis presented in \cite{Bhatt1990} cannot properly deal with probabilistic effects, Bhatt comes close to this interpretation by analysing the kind NP in the GPC as a DP and in the NAC as an NP.}

Put differently, the \NACa\ is a \NACb in disguise, and the \PGCa\ is a \PGCd in disguise.
In a probabilistic framework, the crucial question is what controls speakers' decisions to use a a \NACb in disguise or a \PGCd in disguise.
If the \PGCd\ instantiates the PGC prototype and the \NACb\ the NAC prototype, the choice between \PGCa\ and \NACa\ is expected to be influenced by the degree of similarity to either of these prototypes.
I will argue below that the prototypes differ mainly in that they attract different (classes of) measure nouns more or less strongly, mostly because of degrees of grammaticalisation.
However, if lemmas are attracted with different strengths by the two prototypes, then we can expect a baseline attraction effect.
Nouns which occur more often in the \PGCd\ should favour the \PGCa, and nouns which occur more often in the \NACb\ should favour the \NACa.%
\footnote{Such associations between constructions and words or lemmas are the main object of study in approaches such as Collexeme Analysis \citep{GriesStefanowitsch2004}.}
Put differently, a noun's relative frequencies in the prototypical non-alternating constructions should be indicative of same noun's preference to appear in the two alternating variants.
This baseline attraction effect will be quantified in the corpus study Section~\ref{sec:corpusstudies}.

% GRAMMATICALISATION

@@

% CARDINAL QUANTIFIERS

Furthermore, the grammaticalisation path as described above leads from NPs denoting individuated objects standing in a \textit{part-of} relation to a construction with a more diffuse \textit{quantity-of} relation.
Both types of relations can be numerically quantified -- in as much as a precise number of \textit{parts} or a numerically exact \textit{quantity} can be specified --, but it is much more prototypical of quantities to be specified with numerical precision.
Since the \NACa\ is more closely associated with the \textit{quantity-of} relation, cardinals as attributes of the measure noun are expected to have a higher proportional frequency in the \NACa.
For illustration, (\ref{ex:determiners}) shows the expected variants under this hypothesis.
In (\ref{ex:determiners:nac}), the measure noun is modified by a cardinal \textit{drei} `three', and hence the \NACa\ (kind NP shows case agreement) is preferred.
In (\ref{ex:determiners:pgc}), the measure noun is modified by a non-cardinal determiner \textit{einige} `some', and the \PGCa\ (kind NP is in the genitive) is preferred.

\begin{exe}
  \ex\label{ex:determiners} 
  \begin{xlist}
    \ex{\label{ex:determiners:nac}\gll [[Drei Löffel]\Sub{Nom} [heißer Rum]\Sub{Nom}]\Sub{Nom} sind genug.\\
                                       three spoons hot rum are enough.\\
				     \glt Three spoonful of hot rum are enough.}
    \ex{\label{ex:determiners:pgc}\gll [[Einige Löffel]\Sub{Nom} [heißen Rums]\Sub{Gen}]\Sub{Nom} sind genug\\
                                       some   spoons hot    rum  are   enough.\\
				     \glt A few spoonful of hot rum are enough.}
  \end{xlist}
\end{exe}

The statistical models presented in Section~\ref{sec:corpusstudies} will be specified in a way that they could detect such a preference.

Finally, it has often been iterated that the \PGCa\ is more typical of higher registers or even exclusive to written language (see \citealp[320--323]{Hentschel1993}, where she also refers to descriptive and normative grammars).
This is not surprising in as much as the genitive -- an intrinsic part of the PGC -- is generally underrepresented in colloquial vernacular variants of German as a result of a diachronic process wherein many (but not all) uses of the genitive are replaced by other cases or periphrastic constructions \citep{FleischerSchallert2011}.
Under an integral view of prototypes, which incorporate effects related to larger contexts and registers, such preferences can be to be part of the construction prototypes.
In the corpus study in Section~\ref{sec:corpusstudies}, register effects will therefore be modelled -- even if only using some very simple proxies.

@@

\begin{enumerate}[i.]
  \item The relative frequencies with which measure noun lemmas and kind noun lemmas appear in the prototypical (non-alternating) \PGCd\ and \NACb\ are positively correlated with the relative frequency the alternating \PGCa\ and the \NACa\ are chosen.
  \item (Classes of) more strongly grammaticalised measure nouns favour the NAC.
  \item Measure nouns quantified by cardinals favour the PGC variant.
  \item The NAC is associated more strongly with higher registers.
\end{enumerate}

@@

For the present study, I used the German \textit{Corpus from the Web} (COW) in its 2014 version DECOW14A (\citealp{SchaeferBildhauer2012full,Schaefer2015b} and \citealp{BiemannEa2013,SchaeferBildhauer2013} for overviews of web corpora in general and the methodology of their construction), which contains almost 21 billion tokens.%
\footnote{The corpora are made available for free at \url{https://www.webcorpora.org}.
At the time of this writing, a newer 2016 version DECOW16 had already been released.}
I chose this corpus for two main reasons.
First, the external validity of any study is increased through a higher heterogeneity of the sample \citep[30]{MaxwellDelaney2004}, and the DECOW corpus has clearly a much more heterogeneous composition compared to the only other very large corpus of German, the DeReKo \citep{KupietzEa2010} of the Institute for the German Language (IDS), which contains almost exclusively newspaper texts.%
\footnote{It was shown in \cite{W16-2601} that, for example, the spread of topics is much smaller in DeReKo compared to DECOW.}
Second, it was already mentioned that normative grammars often adopt clear positions regarding the grammaticality of either the \NACa\ or the \PGCa.
Thus, newspaper text or any other text that conforms strongly to normative grammars might not represent the alternation phenomenon fully (and without bias) because authors and proof-readers might favour one alternative or the other.
Web corpora, on the other hand, contain at least some amount of non-standard language from forums and similar sources.
For these or similar reasons, COW corpora have been used in a number of peer-reviewed publications, for example \cite{VanGoethemHiligsmann2014,VanGoethemHuening2015,MuellerS2014,Schaefer2016c,SchaeferSayatz2014,SchaeferSayatz2016,Zimmer2015}. 
Therefore, DECOW can be considered the obvious choice for this study.%
\footnote{The use of web data for linguistic research does require explicit and careful justification.
Due to the noisy nature and unknown composition of the web, only carefully designed and established web corpora like the COW corpora or the SketchEngine corpora \citep{KilgarriffEa2014} should be used.
Clearly, using search engine results is \textit{bad science} for many reasons, most prominently total irreproducibility of results, as \cite{Kilgarriff2006} pointed out more than ten years ago.
Careless use of search engine results is still found, however, see \citet[171--175]{DeclerckBrems2016}.}

@@

I used exactly the same stimuli as in the forced choice experiments.
Each participant read both the $16$ sentences with the variant predicted by the corpus model and the $16$ modified sentences with the variant that the corpus model does not predict.%
\footnote{Notice that lemmas and their frequencies as well as lemma classes are included as regressors in the corpus-based GLMM, and there was consequently no additional controlling of lemma frequencies, etc.}
To minimise repetition effects, the stimuli for each participant were separated into two blocks of $16$ targets and $33$ fillers per block.
In the experiment, participants first read all sentences from the first block, then all sentences from the second block.
It was made sure that from each target sentence pair, one sentence was assigned to the first block, and the other sentence to the other block.
The assignment of members of the individual sentence pairs to the blocks was randomised for each participant individually, and so was the order within each block.
The sentences from each pair of variants were kept as far apart as possible.
The fillers also came in pairs such that the second block exclusively contained sentences to which participants had been exposed in the first block in slightly modified form.
In total, each participant read $98$ sentences.
After each sentence, participants had to answer simple (non-metalinguistic) yes-no questions about the previous sentence as distractors.
The distractor questions were different ones in the first and the second block.
There were $38$ (different) participants recruited in exactly the same manner as for the experiment reported in Section~\ref{sec:exp:fc}.
The experiment was conducted using \textit{PsychoPy}.

The reading times were residualised per speaker based on the reading times of all words (not just the targets) by that speaker.
The adjective and the kind noun (\ie\ the constituents bearing the critical case markers) were used as the target region, such as the bracketed words in the example \textit{zwei Gläser} [\textit{sprudelndes Wasser}] `two glasses of sparkling water'.
Outliers farther than 2 inter-quartile ranges from the mean logarithmised residual time were removed ($64$ data points), resulting in a total number of $n=1,152$ observations.
An LMM was specified with the logarithmised residual reading times as the response variable, and the probabilities derived from the corpus GLMM (\textit{Modelprediction}) as the main regressor.
Since the corpus GLMM predicts the probability of the \PGCa (vs.\ the \NACa), it is expected that reading times are positively correlated (longer reading times) with the probability if the stimulus actually contains the \NACa, and negatively correlated (shorter reading times) if the stimulus actually contains the \PGCa.
To account for this, an interaction between \textit{Modelprediction} and \textit{Construction} (levels \textit{PGCa} and \textit{NACa}) was added to the model.
Furthermore, the position ($1$--$98$) of the sentence in the individual experiment (\textit{Position} was included as a fixed effect to control for increasing reading speed during experiment runs.
Random intercepts were specified for \textit{Participant} and \textit{Item} (the $16$ sentence pairs are one \textit{Item} each).%
\footnote{I tried random slopes in order to keep the random effect structure maximal \citep{BarrEa2013}, but it was impossible to get the algorithm to converge due to the added complexity of the interaction.}

@@

The overall model quality is acceptable, and the effect plot for the main effect is shown in Figure~\ref{fig:spr:effects}.
The estimate for the sentences with \NACa\ is obviously imprecise, although pointing into the right direction (longer reading times when the corpus model predicts the \PGCa).
There is a clearer effect in the sentences with \PGCa, which is also confirmed by the `significant' results from the bootstrapped confidence intervals (see Table~\ref{tab:exp:spr}) and from the PB test reported above.
The \PGCa\ brings about an increased reading time ($\beta_{\text{ConstructionPGCa}}=0.062$), which is plausible because it is the much rarer construction (see Section~\ref{sec:corpusstudies}).
However, if it occurs in a prototypical context and with prototypical lexical material, reading times drop ($\beta_{\text{ConstructionPGCa:Modelprediction}}=-0.132$).
This can be seen in the downward slope of curve in the right panel of Figure~\ref{fig:spr:effects}.
This fits into the general picture in as much as the construction with the lower frequency might be developing towards a more sharply defined prototype.%
\footnote{In this context, it should be remembered from Section~\ref{sec:gettingdata} that even the \PGCd\ is much rarer that the \NACb\ ($17,252$ vs.\ $315,635$ occurrences in the auxiliary corpus samples).}
Conversely, the \NACa\ (like the NAC in general) might be the highly frequent default which does not incur reading time penalty, even if it is not the optimal choice in the given context and with the given lexical material.
This interpretation will be considered, among other things, in the next and final section.
